Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-09-20 manuscript/manuscript.tex
Move the shell expression to an appendix
The derivation is not directly relevant to the error analysis because we
don't mention the expressions in the rest of the text. I moved this to
the appendix so it won't interrupt the flow of the text. Being in the
appendix doesn't make it less important or new. It just serves to make
the text for fluid.

\section{Determination of distance-size and delta ratios}

The discretization algorithms described on the previous section are
introduced to improve the accuracy of the numerical approximation of
the gravitational fields.
They involve two predefined ratios that determine how many
times each tesseroid will be discretized and thus indirectly control
the numerical error of the approximation.
The modified version of the adaptive discretization algorithm involves
the distance-size ratio $D$, while the density-based discretization
algorithm makes use of the delta ratio $\delta$.
These algorithms need predefined default values for $D$ and $\delta$
in order to ensure both acceptable numerical accuracy and computation
effectiveness.

On the homogeneous density tesseroids case, \citet{Uieda2016} compared
their numerical model with the analytical solution of a spherical
shell in order to obtain default values for the distance-size ratio
$D$.
We could follow this idea, but for our needs the spherical shell must
have the same density function as the numerical model.
In contrast with the homogeneous density case, the analytical solutions
for a variable density spherical shell are not present in the
literature, so they must be obtained.

We will then perform these comparisons between the analytical solutions
and the numerical model for specific density functions: a linear and an
exponential one.
From these results we will finally obtain default values for $D$
and $\delta$ that ensure a numerical error lower than 0.1\%.

\subsection{Analytical Solutions for Spherical Shell}

The integrals that define the gravity fields in equations
\ref{eq:tesseroid-pot}-\ref{eq:tesseroid-tensor} have almost trivial
analytical solutions in case of a spherical shell with homogeneous
density \citep{Mikuska2006,Grombein2013}.
But for a variable density spherical shell the problem haven't been
addressed in the literature yet.

Lets consider a spherical shell with inner and outer radii $R_1$ and
$R_2$, respectively, whose density depends on the radial spherical
coordinate (Figure \ref{fig:spherical-shell}).
The gravity potential generated by the shell on an arbitrary external
point $Q(0,0,r)$, located along the $z$ axe at a distance $r$ from the
origin, can be written as follows:

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth]{figures/spherical-shell.pdf}
\caption{
    Spherical shell with inner and outer radii $R_1$ and $R_2$, respectively.
    The computation point $Q$ is located on the $z$ axe at a distance $r$ from
    the center of the shell.
    For our purposes we will assume that $Q$ is outside the outer radii,
    i.e. $r > R_2$.
    }
\label{fig:spherical-shell}
\end{figure}

\begin{equation}
    V_\text{sh}(r) = G
    \int\limits_0^{2\pi}
    \int\limits_{-\frac{\pi}{2}}^\frac{\pi}{2}
    \int\limits_{R_1}^{R_2}
    \frac{\rho(r')}{\ell} {r'}^2 \cos\phi' \,
    dr' d\phi' d\lambda',
\end{equation}

\noindent where $\ell$ is defined in equation \ref{eq:ell}. The computation
point $Q$ is located at latitude $\phi=90^\circ$, so $\ell$ and $\cos\psi$
(defined on equations \ref{eq:ell} and \ref{eq:cospsi}) get simplified:

\begin{equation}
    \cos\psi = \sin\phi', \quad
    \ell = \sqrt{r'^2 + r^2 - 2 r r' \sin\phi'}.
\end{equation}

Due to the rotational symmetry along the $z$ axe, the integration in
$\lambda'$ is straightforward:

For our purposes we can assume that the computation point $Q$ is
farther from the origin than the outer radius, i.e. $r>R_2$.
And taking into account that $r'$ is always lower than $R_2$, we can
simplify the square roots in equation \ref{eq:shell-pot-sqrts}:

\begin{equation}
    \sqrt{r^2 + r'^2 + 2rr'} = |r + r'| = r + r',
\end{equation}
\begin{equation}
    \sqrt{r^2 + r'^2 - 2rr'} = |r - r'| = r - r',
\end{equation}

\noindent what leads to the following expression of the potential:

The equation \ref{eq:shell-pot} allows to easily obtain the exact
gravity potential generated by a spherical shell with a variable
density in depth on any outside point located on the $z$ axe.
Due to the rotational symmetry along any axe that passes through the
center of the shell, the expression in equation \ref{eq:shell-pot}
reproduces the gravity potential field on any outside point at distance
$r$ from its center.

The gradient and the Marussi tensor derived from potentials that
depends solely on $r$ have only a few non zero components: the vertical
component of the gradient ($g_z$) and the diagonal components of the
tensor ($g_{xx}$, $g_{yy}$, $g_{zz}$).
They can be easily obtained as \citet{Grombein2013} does, even for any
arbitrary density function $\rho(r')$:

\begin{equation}
    g_z = \frac{V_\text{sh}(r)}{r},
\end{equation}
\begin{equation}
    g_{xx} = g_{yy} = -\frac{V_\text{sh}(r)}{r^2}, \quad
    g_{zz} = \frac{2V_\text{sh}(r)}{r^2}.
\end{equation}

@@

First, by solving the integral on equation \ref{eq:shell-pot}, we obtained the
gravitational potential generated by a spherical shell with this density on
any external point at distance $r$ from its center:

\noindent The first term on this equation reproduces the potential generated
by a spherical shell with variable density $\rho(r') = ar'$, while the second
one constitutes the potential generated by a spherical shell with homogeneous
density $\rho = b$ \citep{Mikuska2006,Grombein2013}.

@@

>>>>

\section{Determination of the distance-size and delta ratios}

The distance-size ration $D$ of the adaptive discretization and the delta ratio $\delta$
of the density-based discretization determine how many times each tesseroid will be
divided and thus indirectly control the numerical error of the integration.
Optimal values for $D$ and $\delta$ must be determined in order to ensure both
acceptable numerical accuracy and computation efficiency for the algorithm.

\citet{Uieda2016} compared the numerical integration of homogeneous density tesseroids
with the analytical solution of a spherical shell \citep{Mikuska2006,Grombein2013} in
order to obtain default values for the distance-size ratio $D$.
We will follow this idea but for our needs the spherical shell must
have the same density function of radius as our tesseroid model.
Analytical solutions for a variable density spherical shell are not present in the
literature and so they must be obtained first.
We derive the expressions for the gravitational fields of spherical shells with linear
and exponential density functions of radius in Appendix \ref{sec:shell}.

We perform comparisons between the analytical solutions for the spherical shell  and the
numerical integration results for linear and exponential density functions.
From these results, we generalize default values for $D$ and $\delta$ that ensure a
numerical error lower than 0.1\% of the spherical shell values.

@@

@@

\appendix

\section{Analytical Solutions for Spherical Shell}
\label{sec:shell}

\begin{figure}
\centering
\includegraphics[width=0.7\linewidth]{figures/spherical-shell.pdf}
\caption{
    Spherical shell with inner and outer radii $R_1$ and $R_2$, respectively.
    The computation point $Q$ is located on the $z$ axis at a distance $r$ from
    the origin of the coordinates system.
    For our purposes we will assume that $Q$ is outside of the shell,
    i.e. $r > R_2$.
}
\label{fig:spherical-shell}
\end{figure}

Consider a spherical shell with inner radius $R_1$ and outer radius $R_2$,
whose density is a function $\rho(r')$ of the radial coordinate
(Fig. \ref{fig:spherical-shell}).
The gravitational potential generated by the shell on an arbitrary external
point $Q$ can be written as follows:

\begin{equation}
    V_\text{sh}(\phi, \lambda, r) = G
    \int\limits_0^{2\pi}
    \int\limits_{-\frac{\pi}{2}}^\frac{\pi}{2}
    \int\limits_{R_1}^{R_2}
    \frac{\rho(r')}{\ell} {r'}^2 \cos\phi' \,
    dr' d\phi' d\lambda',
\end{equation}

\noindent where $\ell$ is defined in equation \ref{eq:ell}.

For $Q$ located along the $z$ axis (i.e., $\phi=90^\circ$) at a distance $r$ from the
origin, equation \ref{eq:ell} simplifies to:

\begin{equation}
    \ell = \sqrt{r'^2 + r^2 - 2 r r' \sin\phi'}.
\end{equation}

\noindent
Because of the rotational symmetry along the $z$ axe, the integration in $\lambda'$ is
straightforward:

Because the computation point $Q$ is outside of the shell, $r > r'$ and the square roots
in equation \ref{eq:shell-pot-sqrts} simplify to

\begin{equation}
    \sqrt{r^2 + r'^2 + 2rr'} = |r + r'| = r + r',
\end{equation}

\begin{equation}
    \sqrt{r^2 + r'^2 - 2rr'} = |r - r'| = r - r',
\end{equation}

\noindent which leads to the following expression for the potential:

The gradient and the Marussi tensor of potentials that
depends solely on $r$ have only a few non zero components: the vertical
component of the gradient ($g_z$) and the diagonal components of the
tensor ($g_{xx}$, $g_{yy}$, $g_{zz}$).
Following \citet{Grombein2013}:

\begin{equation}
    g_z(r) = \frac{V_\text{sh}(r)}{r},
\end{equation}

\begin{equation}
    g_{xx}(r) = g_{yy}(r) = -\frac{V_\text{sh}(r)}{r^2}
\end{equation}

\begin{equation}
    g_{zz}(r) = \frac{2V_\text{sh}(r)}{r^2}.
\end{equation}

From Eq. \ref{eq:shell-pot} we can obtain expressions for gravitational potential for
different density functions.
For a linear density function

\begin{equation}
    \rho(r') = ar' + b\ ,
\end{equation}

\noindent
the gravitational potential at any external point is

\noindent The first term on this equation reproduces the potential generated
by a spherical shell with variable density $\rho(r') = ar'$, while the second
term constitutes the potential generated by a spherical shell with homogeneous
density $\rho = b$ \citep{Mikuska2006,Grombein2013}.
